---
title: "Midterm 2"
output: html_document
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
path <- "~/Library/CloudStorage/OneDrive-UniversityofCalifornia,Davis/ESP106/esp106-kelsea_macdonald/mid2"
setwd(path)
```

### Data 

For this midterm you need to use two datasets:

"chinalanduse_MODIS_2012.nc" contains four layers with land cover data for China. The data were derived from MODIS satellite data for the year 2012. Each layer gives the fraction of the grid cell that has a specific land cover type: urban (layer 1), cropland (layer 2), grassland (layer 3) and forest (layer 4). 

"ch_adm.*" with polygons for the provinces of China.


#### Q1. Read in the land use data as a SpatRaster get the polygons as a SpatVector (2 points)


```{r}
file.names <- list.files(path)
library(terra)
land_use <- rast("chinalanduse_MODIS_2012.nc")
provinces <- vect("chn_adm.shp")
provinces <- subset(provinces, provinces$ENGTYPE_1 == "Province")
```

#### Q2a. Crop the land use SpatRaster to the same extent as the SpatVector of Chinese provinces (1 point), and set all grid cells outside of China to `NA`

```{r}
land_use <- crop(land_use, ext(provinces))
land_use <- mask(land_use, provinces, updatevalue=NA)
```

#### Q2b. Rename the layers in the SpatRaster so they provide information about what data is in each of the 4 layers (2 points)

```{r}
names(land_use) <- c("urban", "cropland", "grassland", "forest")
```

#### Q3. Make a figure showing each SpatRaster layer on one of the panels and overlay the polygons of the Chinese provinces. Title each panel with the type of land use it shows. (4 points)

```{r}
par(mfrow = c(2,2))
plot(land_use)
# plot(provinces, add=TRUE)
# plot(provinces)
# I could not get this province polygons to plot on top of the raster data
```

#### Q4a. Use `terra::extract` to find the fraction of each province in each of the four land use classes. [For this question you can assume all the grid cells have the same size] (3 points)

```{r}
tab <- terra::extract(land_use, provinces)
#I assigned each ID its unique province name so the table is more meaningful
names <- data.frame("ID" = c(1:22), "province" = unique(provinces$NAME_1))
tab <- merge(names, tab, by = "ID")
# tab = the data.frame of land-use fractions by province for each grid cell

fractions <- aggregate(cbind(urban,cropland, grassland, forest) ~ 
                         province, data = tab, mean)
# fractions = the data.frame of land-use fractions averaged for each province
fractions
```

#### Q4b. Describe the potential problem with the area assumption made in 4a. How might it affect the calculation in that step? What could we do if we didn't want to make that assumption? (You don't have to do it, just describe in theory) (2 points)

**Answer: We assume all grid cells are of equal size, but in reality we put a mask on the raster so the edge-cells are not full squares. This means we are calculating for more data than actually exists within the Chinese borders. If we didn't want to make this assumption we could take the fraction of the area that the border cells do make up and multiply it by our land-use data to get the actual values.**


#### Q4c. Sum up the fractions in the four land cover classes for each province and plot these as a histogram. (2 points) 

```{r}
# I used rowSums to get the sum of all the fractions for each province, excluding the ID column
summed.fractions <- data.frame("province" = fractions$province, 
                               "sum" = rowSums(fractions[,-1]))

par(mfrow = c(1,1))
hist(summed.fractions[,2], 
     main = "frequency of total land cover fractions for 
     \n22 provinces in China")
```


#### Q5. Add a new variable "other" to the data created with terra::extract, and give it the appropriate value. (2 points)

```{r}
fractions <- cbind("ID" = c(1:22), fractions, "other" = 1-summed.fractions$sum)
#This creates a new column with the remaining percent from the totals
```


#### Q6. Make barplots showing the breakdown of urban, cropland, grassland, forest, and other for each province. The barplots should be "stacked" (a single bar for each province, showing land cover with a color) and "horizontal" (province names on the vertical axis).  

#### Q6a) Use graphics::barplot, make sure to include a legend.  (4 points)

```{r}
matrix.fractions <- as.matrix(fractions[,c(1,3:7)])
tmatrix <- t(matrix.fractions) 
#I flipped the orientation of my matrix in order to get the stacking I wanted

graphics::barplot(height = tmatrix[2:6,], horiz = TRUE,
                  names.arg = fractions$province, cex.names = .5, las = 1,
                  col = c("blue", "yellow", "green", "darkgreen", "grey"),
                  main = "Land Use in China")
legend(x = "bottomright", legend = c("urban", "cropland", "grassland",
                  "forest", "other"), col = c("blue","yellow","green",
                  "darkgreen","grey"), pch=15, cex=0.8, pt.cex=0.8)
```

#### Q6b) Use ggplot. (4 points) 

```{r}
library(ggplot2)
# First I need to rearrange my data.frame so it is easier for ggplot2 to read
# I used a method I found online which turns each land use type into a "group" value which I can use as an identity statistic for filling the bars
fractions.flipped <- data.frame(province = fractions$province,
                                percent = c(fractions$urban,
                                            fractions$cropland,
                                            fractions$grassland, 
                                            fractions$forest,
                                            fractions$other),
                       group = c(rep("urban", nrow(fractions)),
                                 rep("cropland", nrow(fractions)),
                                 rep("grassland", nrow(fractions)),
                                 rep("forest", nrow(fractions)),
                                 rep("other", nrow(fractions))))
ggplot(fractions.flipped, aes(province, percent, fill = group)) +
  geom_bar(stat="identity")+
  coord_flip()
```


#### Q7. Upload your R markdown file, and your knitted output to Canvas. Push the R markdown file to your Github repository. (2 points)

